---
title: "GPP model for NHC"
author: "Alice Carter"
date: "2/15/2022"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = 'C:/Users/Alicei Carter/git/nhc_50yl')
setwd('C:/Users/Alice Carter/git/nhc_50yl')
source('../autotrophic_rivers/src/stan_helpers.R')
library(tidyverse)
library(rstan)
```
Task: build a basic hierarchical model of productivity as a function of light, temperature, and discharge.

### Initial model

Model productivity $P$ as a function of an underlying biomass term $(B)$ with light $(L)$ and temperature $(T)$ dependent photosynthetic rates and a biomass loss determined by discharge $(Q)$.

$$P_t = P_{20} \times L_t e^{B_t} \times e^{(\frac{E}{k_b} \times (\frac{1}{293} - \frac{1}{T_t}))}$$
$$B_t = (B_{t-1} + r + \lambda e^{B_{t-1}}) \times (1 - \beta Q_t) + \epsilon_t$$
$$\epsilon_t \sim N(0,\sigma_{proc})$$
$$P_{obs,t} \sim N(P_t, \sigma_{obs})$$
Simulate some data based on real driver variables:

```{r}

dat <- read_csv("data/metabolism/metabolism_and_drivers.csv")
drivers <- dat %>%
  filter(site == 'CBP') %>%
  mutate(GPP_sd = (GPP.upper - GPP.lower)*.33/.95) %>%
  select(date, PAR_surface, LAI, temp.water, discharge, GPP, GPP_sd) %>%
  mutate(across(-date, zoo::na.approx, na.rm = F),
         logQ = (log(discharge)),
         tQ = discharge/max(discharge, na.rm = T),
         light = PAR_surface/max(PAR_surface, na.rm = T),
         tempK = temp.water+ 273) %>%
  arrange(date) %>%
  group_by(date) %>%
  summarize(across(everything(), mean, na.rm = T)) %>%
  mutate(across(any_of(c('tempK', 'tQ', 'light')), zoo::na.approx, na.rm = F)) %>%
  slice(-1)

beta = 0.8
r = .06
lambda = -.01
sig_obs = 0.05
sig_proc = 0.05

E = 0.35
k_b = 8.6173 * 10 ^ -5
P_20 = 1

B = numeric()
pred_P = numeric()

B[1] = rnorm(1, log(drivers$GPP[1]/drivers$light[1]), 1)
pred_P[1] = drivers$light[1] * exp(B[1]) * P_20 * exp(E/k_b *( 1/293 - 1/drivers$tempK[1]))

N = nrow(drivers)

for (i in 2:N){
  B[i] = rnorm(1, ((B[i-1] + r + lambda * exp(B[i-1])) * (1-beta*drivers$tQ[i])), sig_proc)
  pred_P[i] = drivers$light[i] * exp(B[i]) * P_20 * exp(E/k_b *(1/293 - 1/drivers$tempK[i]))
}

P_obs <- rnorm(N, pred_P, sig_obs)

plot(drivers$date, P_obs)
lines(drivers$date, drivers$GPP)


plot(drivers$date, B)
```


## Stan code

```{r, comment='', echo = FALSE}
cat(read_lines('src/model/stan/ss_GPP.stan'), sep = '\n')
cat(read_lines('src/model/stan/Stan_ProductivityModel2_Ricker_fixedinit_obserr.stan'), sep = '\n')
```

```{r}
stan_dat <- list(N = nrow(drivers), P_obs = P_obs, light = drivers$light,
                 tempK = drivers$tempK, Q = drivers$tQ)
stan_dat <- list(N = nrow(drivers), P_obs = drivers$GPP, light = drivers$light,
                 tempK = drivers$tempK, Q = drivers$tQ)

mod <- stan('src/model/stan/ss_GPP.stan', 
            data = stan_dat, init = 0,
            chains = 1, cores = 4, 
            warmup = 200, iter = 500)

# try running Heili's ricker model

stan_dat <- list(Ndays = nrow(drivers), light = drivers$light, GPP = drivers$GPP, 
                 GPP_sd = drivers$GPP_sd,
                 tQ = drivers$tQ)

  mod <- stan('src/model/stan/Stan_ProductivityModel2_Ricker_fixedinit_obserr.stan', 
            data = stan_dat,# init = 0,
            chains = 4, cores = 4, 
            warmup = 200, iter = 500)

```

```{r, echo = FALSE}
saveRDS(mod, 'src/model/model_runs/simulated_ss_GPP.rds')
mod <- readRDS('src/model/model_runs/simulated_ss_GPP.rds')
print(mod, pars = c( 'r','lambda',  'sig_p', 'sig_o'))
print(mod, pars = c('beta', 'r','lambda',  'P_20', 'sigma_proc', 'sigma_obs'))
traceplot(mod, ncol = 1, pars=c('phi','b_light', 'b_q', 'sigma_obs', 'sigma_proc'))
plot_post_sim(mod, pars = c('phi', 'b_light', 'b_q', 'sigma_obs', 'sigma_proc'), 
              vals = c(phi, b_light, b_q, sig_obs, sig_proc))
```

This model is not doing a good job of recovering the $\beta_q$ parameter

### Latent biomass model

An alternative approach to the phenomenological one presented above is a model that takes into account the latent biomass of primary producers to model $P$ mechanistically. Here, I am implementing a version of this type of model from Segatto et al 2020. I am choosing to leave out the density dependance of Biomass, because I do not think that NHC is ever space limited.

$$P(t) = \mu_{P,20} f_{T,L}(t)B_A(t)$$
$$f_{T,L}(t) = \theta_A^{T(t)-20} \frac{PAR(t)}{PAR(t) + K_{PAR}$$
$$B_A(t) = B_A(t-1) + P(t-1) - e^{s(Q(t) - c)}$$
ary(tidyverse)
Where $\
library(lme4)

setwd('C:/Users/Alice Carter/git/nhc_50yl/')

sites <- read_csv('data/siteData/NHCsite_metadata.csv') %>%
  slice(c(1:5, 7))

# Test basic GPP model
P <- dat %>% 
  select(date, site, GPP, GPP.lower, GPP.upper, discharge, 
         temp.water, light, slope) %>%
  group_by(site) %>%
  mutate(GPP_pre = c(NA, GPP[1:(n()-1)]))

P <- filter(P, site == 'CBP') %>%
  mutate(discharge = log(discharge),
         across(.cols = any_of(c('light', 'discharge')), .fns = scale))
mod <- lm(GPP ~ GPP_pre + light + discharge, data = P)
summary(mod)

phi = 0.83
b_light = 0.0364
b_q = .06826
p = numeric()
p[1] = P$GPP[2]
for(i in 2:nrow(P)){
  p[i] = phi * p[i-1] + b_light * P$light[i] + b_q * P$discharge[i]
}
plot(P$date, p, ylim = c(-1,2))
lines(P$date, P$GPP)
plot(P$date, P$light)
m = T))

shapiro.test(dat$GPP)
library(ggpubr)
ggqqplot(dat$ER)
aes(temp.water, GPP, col = factor(month))) +
  geom_point() +
  facet_wrap(~site, scales = 'free_y') +

  geom_smooth(method = lm)

mod <- brms::brm(ER ~ (1 + temp.water| slope +  logQ_mean), data = dat, iter = 1000)
mod


raopagaagLFMrcssResnittlstceua`td%=dae_am