---
title: "GPP model for NHC"
author: "Alice Carter"
date: "2/15/2022"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = 'C:/Users/Alice Carter/git/nhc_50yl')
setwd('C:/Users/Alice Carter/git/nhc_50yl')
library(tidyverse)
library(rstan)
```
Task: build a basic hierarchical model of productivity as a function of light, temperature, and discharge.

### Initial model

To start, I will model productivity $P$ as an autoregressive function with terms for light ($l$), temp ($T$), and discharge ($Q$) included. If this proves insufficient, I will incorporate a latent biomass term.

$$P_t = (\phi P_{t-1} + \beta_l l_t) \times (1 - \beta_q Q_t) \times e^{(\frac{1}{T_t} - \frac{1}{293})} + \epsilon_t$$
$$\epsilon_t \sim N(0,\sigma_{proc})$$
$$P_{obs,t} \sim N(P_t, \sigma_{obs})$$
Simulate some data based on real driver variables:

```{r}

dat <- read_csv("data/metabolism/metabolism_and_drivers.csv")
drivers <- dat %>%
  filter(site == 'CBP') %>%
  mutate(GPP_sd = (GPP.upper - GPP.lower)*.33/.95) %>%
  select(date, PAR_surface, LAI, temp.water, discharge, GPP, GPP_sd) %>%
  mutate(across(-date, zoo::na.approx, na.rm = F),
         logQ = (log(discharge)),
         tQ = discharge/max(discharge, na.rm = T),
         light = PAR_surface/max(PAR_surface, na.rm = T),
         tempK = temp.water+ 273) %>%
  arrange(date) %>%
  group_by(date) %>%
  summarize(across(everything(), mean, na.rm = T)) %>%
  mutate(across(any_of(c('tempK', 'tQ', 'light')), zoo::na.approx, na.rm = F)) %>%
  slice(-1)

phi = 0.9
b_light = 0.1
b_q = 0.8
sig_obs = 0.05
sig_proc = 0.05

Phat = P = numeric()
Phat[1] = P[1] = .5
N = nrow(drivers)

for (i in 2:N){
  Phat[i] = (phi *P[i-1] + b_light * drivers$light[i])*(1-b_q*drivers$tQ)*
    exp(1/293 - 1/drivers$tempK) 
  P[i] = rnorm(1, Phat[i], sig_proc)
  if(P[i] < 0) P[i] = 0
}

P_obs <- rnorm(N, P, sig_obs)

plot(drivers$date, P_obs, ylim = c(0, 2))
lines(drivers$date, drivers$GPP)

```


## Stan code

```{r, comment='', echo = FALSE}
cat(read_lines('src/model/stan/ss_GPP.stan'), sep = '\n')
```

```{r}
stan_dat <- list(N = length(P_obs), P_obs = P_obs, light = drivers$light,
                 tempK = drivers$tempK, Q = drivers$tQ)

mod <- stan('src/model/stan/ss_GPP.stan', 
            data = stan_dat, #init = 0,
            chains = 4, cores = 4, 
            warmup = 200, iter = 500)

```

```{r, echo = FALSE}
# saveRDS(mod, 'src/model/model_runs/simulated_ss_GPP.rds')
mod <- readRDS('src/model/model_runs/simulated_ss_GPP.rds')
print(mod, pars = c('phi', 'b_light','b_q',  'sigma_obs', 'sigma_proc'))
traceplot(mod, ncol = 1, pars=c('phi','b_light', 'b_q', 'sigma_obs', 'sigma_proc'))
plot_post_sim(mod, pars = c('phi', 'b_light', 'b_q', 'sigma_obs', 'sigma_proc'), 
              vals = c(phi, b_light, b_q, sig_obs, sig_proc))
```

This model is not doing a good job of recovering the $\beta_q$ parameter

### Latent biomass model

An alternative approach to the phenomenological one presented above is a model that takes into account the latent biomass of primary producers to model $P$ mechanistically. Here, I am implementing a version of this type of model from Segatto et al 2020. I am choosing to leave out the density dependance of Biomass, because I do not think that NHC is ever space limited.

$$P(t) = \mu_{P,20} f_{T,L}(t)B_A(t)$$
$$f_{T,L}(t) = \theta_A^{T(t)-20} \frac{PAR(t)}{PAR(t) + K_{PAR}$$
$$B_A(t) = B_A(t-1) + P(t-1) - e^{s(Q(t) - c)}$$
ary(tidyverse)
Where $\
library(lme4)

setwd('C:/Users/Alice Carter/git/nhc_50yl/')

sites <- read_csv('data/siteData/NHCsite_metadata.csv') %>%
  slice(c(1:5, 7))

# Test basic GPP model
P <- dat %>% 
  select(date, site, GPP, GPP.lower, GPP.upper, discharge, 
         temp.water, light, slope) %>%
  group_by(site) %>%
  mutate(GPP_pre = c(NA, GPP[1:(n()-1)]))

P <- filter(P, site == 'CBP') %>%
  mutate(discharge = log(discharge),
         across(.cols = any_of(c('light', 'discharge')), .fns = scale))
mod <- lm(GPP ~ GPP_pre + light + discharge, data = P)
summary(mod)

phi = 0.83
b_light = 0.0364
b_q = .06826
p = numeric()
p[1] = P$GPP[2]
for(i in 2:nrow(P)){
  p[i] = phi * p[i-1] + b_light * P$light[i] + b_q * P$discharge[i]
}
plot(P$date, p, ylim = c(-1,2))
lines(P$date, P$GPP)
plot(P$date, P$light)
m = T))

shapiro.test(dat$GPP)
library(ggpubr)
ggqqplot(dat$ER)
aes(temp.water, GPP, col = factor(month))) +
  geom_point() +
  facet_wrap(~site, scales = 'free_y') +

  geom_smooth(method = lm)

mod <- brms::brm(ER ~ (1 + temp.water| slope +  logQ_mean), data = dat, iter = 1000)
mod


raopagaagLFMrcssResnittlstceua`td%=dae_am